---
title: "Air Quality Analysis"
author: "Aditya Chauhan"
date: "`r Sys.Date()`"
output: 
  pdf_document:
    fig_width: 12
    fig_height: 8
    fig_caption: true
header-includes:
   - \usepackage{geometry}
   - \geometry{margin=1in}
   - \usepackage{float}
   - \floatplacement{figure}{H}
---


---------------------------------

### Name: Aditya Chauhan

### ID: 169027493

### Class: DATA 100 A

### Professor: Dr. Devon Becker

---------------------------------

# Introduction

## Dataset

Air Quality (UC Irvine) (Vito, S. (2008). Air Quality [Dataset]. UCI Machine Learning Repository. <https://doi.org/10.24432/C59K5F>.)

## Abstract

Accurate air quality monitoring relies on sensor data to measure pollutants such as ozone, carbon monoxide, and nitrogen oxides. This study examines the correlation between sensor readings and measured pollutant levels to validate monitoring technologies and enhance their predictive capabilities. By leveraging models, I aim to predict carbon monoxide concentrations based on sensor data and other environmental factors.

# Goals/Research Question

Descriptive Features:
- Continuous
  - Sensor targeting CO - PT08.S1(CO)
  - Sensor targeting O3 - PT08.S5(O3)
  - True averaged sensor response - NOx(GT)
- Categorical
  - Season - Based on Dates
  
Target Feature:
- True Concentration of CO - CO(GT)

# Basic Data Cleaning & Preprocessing

## Library Loads & Data Imports
```{r}
library(readxl)
library(tidyverse)
library(ggplot2)
library(tidymodels)
library(patchwork)
library(lubridate)
library(zoo)
library(reshape2)
library(dplyr)
library(RColorBrewer)
library(ggridges)
library(ggrepel)

air_quality_data <- read_excel("AirQualityUCI.xlsx", sheet = "AirQualityUCI")
```

## Creating Categorical Features, Other Preprocessing, & Advanced Data Cleaning
```{r}

# Data Preprocessing Pipeline
# Steps:
# 1. Create seasonal categories
# 2. Handle missing values and outliers
# 3. Create interaction terms
# 4. Add lagged features

# Define pollutant columns for handling outliers
pollutant_columns <- c("CO(GT)", "NOx(GT)", "NO2(GT)")

# Data Preprocessing Pipeline
air_quality_data <- air_quality_data |>
  mutate(
    Season = case_when(
      format(Date, "%m") %in% c("12", "01", "02") ~ "Winter",
      format(Date, "%m") %in% c("03", "04", "05") ~ "Spring",
      format(Date, "%m") %in% c("06", "07", "08") ~ "Summer",
      format(Date, "%m") %in% c("09", "10", "11") ~ "Autumn"
    )
  ) |>
  
  # This code drops a bunch of values that were all at -200, likely sensor failures since CO & other levels can't be negative or -200 
  drop_na(`CO(GT)`, `NOx(GT)`, `NO2(GT)`) |>
  filter(`CO(GT)` >= 0, `NOx(GT)` >= 0, `NO2(GT)` >= 0) |>
  group_by(Season) |>
  
  # Winsorize outliers
  mutate(
    across(all_of(pollutant_columns),
           ~ ifelse(. > quantile(., 0.95, na.rm = TRUE), quantile(., 0.95, na.rm = TRUE), .))
  ) |>
  ungroup() |>
  
  # Handle missing values using conditional replacement
  mutate(
    across(c(`NOx(GT)`, `NO2(GT)`), ~ na_if(., -200)),
    `NOx(GT)` = if_else(is.na(`NOx(GT)`), mean(`NOx(GT)`, na.rm = TRUE), `NOx(GT)`),
    `NO2(GT)` = if_else(is.na(`NO2(GT)`), mean(`NO2(GT)`, na.rm = TRUE), `NO2(GT)`)
  ) |>
  
  # Standardize Continuous Variables
  mutate(
    across(c(T, RH, AH, `PT08.S1(CO)`, `PT08.S5(O3)`), ~ scale(.))
  ) |>
  
  # Add interaction terms
  mutate(
    NOx_PT08_S1_Interaction = `NOx(GT)` * `PT08.S1(CO)`,
    NOx_Season_Interaction = as.numeric(as.factor(Season)) * `NOx(GT)`
  )

# Create lagged feature for CO(GT) (previous hour)
air_quality_data <- air_quality_data |>
  arrange(Date, Time) |>
  mutate(
    Lagged_CO = lag(`CO(GT)`, n = 1, default = NA)
  )

# Drop rows with missing values introduced by lagged features
air_quality_data <- drop_na(air_quality_data)

# Improved data cleaning for CO(GT)
air_quality_clean <- air_quality_data |>
  # Replace -200 values with NA since they appear to be missing data
  mutate(
    `CO(GT)` = ifelse(`CO(GT)` == -200, NA, `CO(GT)`)
  ) |>
  drop_na(`CO(GT)`) |>
  filter(`CO(GT)` >= 0) 

# Add a custom function for data cleaning
clean_sensor_data <- function(data, threshold) {
  data |>
    mutate(sensor_status = case_when(
      `PT08.S1(CO)` > threshold ~ "High",
      `PT08.S1(CO)` < -threshold ~ "Low",
      TRUE ~ "Normal"
    ))
}

# Add pivot operations for sensor readings
sensor_long <- air_quality_data |>
  pivot_longer(
    cols = starts_with("PT08"),
    names_to = "sensor_type",
    values_to = "reading"
  )

# Final Dataset Split
set.seed(123)
data_split <- initial_split(air_quality_data, prop = 0.6)
train_data <- training(data_split)
temp_data <- testing(data_split)
validation_split <- initial_split(temp_data, prop = 0.5)
validation_data <- training(validation_split)
test_data <- testing(validation_split)
```

# Exploratory & Model Plots

## Exploratory Plot/Table 1
```{r}
plot1 <- ggplot(air_quality_data, aes(x = `NOx(GT)`, y = `CO(GT)`)) +
  geom_hex(bins = 30) +
  scale_fill_viridis_c() +
  geom_smooth(method = "lm", se = TRUE, color = "darkred", linetype = "dashed", size = 1) +
  scale_y_continuous(limits = c(-10, 10), breaks = seq(-10, 10, by = 5)) +
  scale_x_continuous(limits = c(0, 600), breaks = seq(0, 600, by = 100)) +
  labs(
    title = "Density of NOx and CO Measurements",
    x = "NOx(GT) (Nitrogen Oxides)",
    y = "CO(GT) (Carbon Monoxide)",
    fill = "Count"
  ) +
  theme_minimal(base_size = 12) +
  theme(plot.title = element_text(face = "bold"))

# This plot showcases the relationship betwen NOx(GT) and CO(GT), critical to the research question because it visually establishes whether there is a relationship between these variables. It also shows density at certain values. 

```

## Exploratory Plot/Table 2
```{r}
plot2 <- ggplot(air_quality_data, aes(x = `NOx(GT)` * `PT08.S1(CO)`, y = `CO(GT)`, color = Season)) +
  geom_point(alpha = 0.6, size = 1.5) +
  geom_smooth(method = "lm", se = TRUE, color = "orange", linetype = "dotted", size = 1) +
  scale_y_continuous(limits = c(-10, 10), breaks = seq(-10, 10, by = 5)) +
  scale_x_continuous(limits = c(-2000, 1000), breaks = seq(-2000, 1000, by = 500)) +
  scale_color_brewer(palette = "Set2") +
  labs(
    title = "Interaction Between NOx(GT) and Sensor Data on CO(GT)",
    x = "NOx(GT) * PT08.S1(CO) (Interaction Term)",
    y = "CO(GT) (Carbon Monoxide)",
    color = "Season"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title = element_text(face = "bold"),
    legend.position = "bottom"
  )

# This plot examine show the interaction of NOx(GT) and PT08.S1(CO) relates to CO(GT). By categorizing the data by season, it integrates continuous & categorical features and helps with understanding how sensor data correlates with pollutants in different contexts of environment
```

```{r, fig.width=12, fig.height=8, out.width="100%"}
# Combine the two plots
final_exploratory_plot <- plot1 | plot2 +
  plot_layout(guides = "collect") & 
  theme(legend.position = "bottom")
final_exploratory_plot

# In plot 1, the positive correlation between NOx(GT) and CO(GT) indicates that higher NOx levels are associated with higher CO levels, consistent with established emission patterns in urban areas. This aligns with the role of shared sources like vehicle emissions and industrial activities.

# In plot 2, the interaction term highlights the complex relationship between NOx(GT) and PT08.S1(CO), revealing how sensor efficiency varies by season. In winter, the interaction term is often negative, likely due to reduced sensor performance caused by cold temperatures and atmospheric conditions. In contrast, spring and autumn show predominantly positive interaction values, reflecting improved sensor performance, while summer demonstrates the highest efficiency due to favorable environmental factors for pollutant detection.
```

## Model Plot 1
```{r}
model_plot1 <- ggplot(train_data, aes(x = `CO(GT)`, y = Season, fill = Season)) +
  geom_density_ridges(alpha = 0.7) +
  facet_wrap(~ cut(`PT08.S1(CO)`, breaks = 3, labels = c("Low", "Medium", "High"))) +
  geom_label(
    data = train_data |>
      group_by(Season) |> 
      summarise(mean_co = mean(`CO(GT)`), y = as.numeric(factor(first(Season)))),
    aes(x = mean_co, y = y, label = round(mean_co, 2)),
    inherit.aes = FALSE
  ) +
  labs(
    title = "CO Distribution by Season and Sensor Level",
    subtitle = "Seasonal Variations in CO Levels Across Sensor Categories",
    caption = "Note: Sensor levels determined by tertile splits of PT08.S1(CO) readings",
    x = "CO(GT) Ground Truth",
    y = "Season"
  ) +
  theme_minimal()
```

## Model Plot 2
```{r}
model_plot2 <- ggplot(train_data, aes(x = `PT08.S5(O3)`, y = `CO(GT)`, color = `NOx(GT)`)) +
  geom_point(alpha = 0.6) +
  scale_color_viridis_c() +
  geom_smooth(method = "lm", se = TRUE) +
  labs(
    title = "O3 Sensor vs CO with NOx Levels",
    x = "PT08.S5(O3) Sensor Reading (Standardized)", 
    y = "CO(GT) Ground Truth"
  ) +
  theme_minimal()

```

```{r, fig.width=12, fig.height=8, out.width="100%"}
model_plot1 | model_plot2

# The ridgeline plot in Model Plot 1 demonstrates the relationship between multiple features (Seasons, PT08.S1(CO) sensor readings), by showing how CO(GT) distributions vary across different sensor level categories & seasons. This directly address the question about sensor measurement correlations because the distinct peaks and overlapping distributions reveal seasonal patterns in sensor reliability, which clearly varies. 

# The scatter plot in Model Plto 2 examines the relationship between the O3 sensor and CO ground truthm with NOx levels shown through the colour gradient. This is relevant because it shows potential cross-sensitivity between O3 & CO sensors while accounting for NOx interference. 
```

# Linear Models

## Exploratory Linear Model 1
```{r}
# First model focusing on main effects and seasonal interaction
model_with_feature <- linear_reg() |>
  set_engine("lm") |>
  fit(`CO(GT)` ~ `PT08.S1(CO)` + `NOx(GT)` + Season, data = train_data)

model_without_feature <- linear_reg() |>
  set_engine("lm") |>
  fit(`CO(GT)` ~ `PT08.S1(CO)` + Season, data = train_data)

# Calculate RMSE comparison using validation data
val_metrics_with <- augment(model_with_feature, new_data = validation_data) |>
  summarize(rmse = sqrt(mean((`CO(GT)` - .pred)^2)))

val_metrics_without <- augment(model_without_feature, new_data = validation_data) |>
  summarize(rmse = sqrt(mean((`CO(GT)` - .pred)^2)))

# Print results
print("RMSE with NOx(GT):")
print(val_metrics_with)
print("RMSE without NOx(GT):")
print(val_metrics_without)
```

## Exploratory linear model 2
```{r}
# Second model incorporating O3 sensor and interaction terms
model_with_o3 <- linear_reg() |>
  set_engine("lm") |>
  fit(`CO(GT)` ~ `PT08.S1(CO)` + `PT08.S5(O3)` + `NOx(GT)` + Season, data = train_data)

# Add a categorical feature interaction model
model_with_interaction <- linear_reg() |>
  set_engine("lm") |>
  fit(`CO(GT)` ~ `PT08.S1(CO)` * Season + `NOx(GT)`, data = train_data)

# Calculate RMSE for models
val_metrics_interaction <- augment(model_with_interaction, new_data = validation_data) |>
  summarize(rmse = sqrt(mean((`CO(GT)` - .pred)^2)))

val_metrics_o3 <- augment(model_with_o3, new_data = validation_data) |>
  summarize(rmse = sqrt(mean((`CO(GT)` - .pred)^2)))

# Compare all models
model_comparison <- tibble(
  Model = c("With NOx", "Without NOx", "With Interaction", "With O3"),
  RMSE = c(
    val_metrics_with$rmse,
    val_metrics_without$rmse,
    val_metrics_interaction$rmse,
    val_metrics_o3$rmse
  )
) |>
  arrange(RMSE)

```

```{r, fig.width=12, fig.height=8, out.width="100%"}
# Create comparison plot
model_comparison_plot <- ggplot(model_comparison, aes(x = reorder(Model, -RMSE), y = RMSE)) +
  geom_col(aes(fill = RMSE)) +
  scale_fill_viridis_c(direction = -1) +
  geom_text(aes(label = sprintf("%.3f", RMSE)), vjust = -0.5) +
  labs(
    title = "Comparison of Model Performance",
    subtitle = "Model with Seasonal Interactions Performs Best",
    x = "Model Specification",
    y = "Root Mean Square Error (RMSE)",
    caption = "Based on validation set performance\nLower RMSE indicates better predictive accuracy"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    plot.title = element_text(face = "bold"),
    legend.position = "none"
  )

# Display the comparison
print(model_comparison)
model_comparison_plot
```

## Bar Plot
```{r, fig.width=12, fig.height=8, out.width="100%"}
# Create seasonal average bar plot
seasonal_plot <- ggplot(air_quality_data |>
                         group_by(Season) |>
                         summarise(avg_co = mean(`CO(GT)`),
                                 sd_co = sd(`CO(GT)`)), 
                       aes(x = Season, y = avg_co)) +
  geom_col(aes(fill = Season)) +
  geom_errorbar(aes(ymin = avg_co - sd_co, 
                    ymax = avg_co + sd_co), 
                width = 0.2) +
  labs(
    title = "Average CO Levels by Season",
    subtitle = "With Standard Deviation Error Bars",
    x = "Season",
    y = "Average CO(GT)",
    caption = "Error bars represent ±1 standard deviation"
  ) +
  theme_minimal()

# Add this to your plot display section
seasonal_plot
```


# Final Linear Model, Diagnostics & Interpretations
```{r, fig.width=12, fig.height=8, out.width="100%"}
# Train final model on combined train + validation data
final_training_data <- bind_rows(train_data, validation_data)

# Final model with interactions
final_model <- linear_reg() |>
  set_engine("lm") |>
  fit(`CO(GT)` ~ `PT08.S1(CO)` * Season + `NOx(GT)` + `PT08.S5(O3)`, 
      data = final_training_data)

# Evaluate on test set 
final_metrics <- augment(final_model, new_data = test_data) |>
  summarize(
    rmse = sqrt(mean((`CO(GT)` - .pred)^2)),
    r2 = cor(`CO(GT)`, .pred)^2
  )

# Scatter for predicted vs actual
diagnostics_plot1 <- augment(final_model, new_data = test_data) |>
  ggplot(aes(x = .pred, y = `CO(GT)`)) +
  geom_point(alpha = 0.5) +
  geom_abline(color = "red", linetype = "dashed") +
  geom_label_repel(
    data = function(x) {
      x |> 
        filter(abs(.resid) > quantile(abs(.resid), 0.99)) |>
        slice_max(order_by = abs(.resid), n = 3)
    },
    aes(label = round(.resid, 2)),
    box.padding = 0.5,
    max.overlaps = 3
  ) +
  labs(
    title = "Predicted vs Actual CO(GT) Values",
    subtitle = "Test Set Performance",
    x = "Predicted Values",
    y = "Actual Values",
    caption = "Red line indicates perfect prediction. Labels show largest residuals."
  ) +
  theme_minimal()

# Q-Q plot with enhancements
diagnostics_plot2 <- augment(final_model, new_data = test_data) |>
  ggplot(aes(sample = .resid)) +
  stat_qq() +
  stat_qq_line(color = "red") +
  labs(
    title = "Normal Q-Q Plot of Residuals",
    subtitle = "Assessing Normality of Residuals",
    x = "Theoretical Quantiles",
    y = "Sample Quantiles",
    caption = "Labels show extreme residuals"
  ) +
  theme_minimal()

# Residuals vs Fitted plot
diagnostics_plot3 <- augment(final_model, new_data = test_data) |>
  ggplot(aes(x = .pred, y = .resid)) +
  geom_point(alpha = 0.5, aes(color = abs(.resid))) +
  scale_color_viridis_c() +
  geom_hline(yintercept = 0, color = "red", linetype = "dashed") +
  geom_smooth(se = TRUE, color = "blue") +
  labs(
    title = "Residuals vs Fitted Values",
    subtitle = "Checking Homoscedasticity & Linearity",
    x = "Fitted Values",
    y = "Residuals",
    color = "Absolute\nResidual"
  ) +
  theme_minimal() +
  theme(legend.position = "right")

# Updated plot combination
final_diagnostics <- (diagnostics_plot1 | diagnostics_plot2) / diagnostics_plot3 +
  plot_layout(heights = c(2, 1.2)) +
  plot_annotation(
    title = "Model Diagnostic Plots",
    subtitle = sprintf("Final Model RMSE: %.3f | R²: %.3f | Performance on Test Set", 
                      final_metrics$rmse, final_metrics$r2)
  ) &
  theme(plot.title = element_text(face = "bold"))

# Print model summary and diagnostics
print("Final Model Summary:")
tidy(final_model) |>
  mutate(across(where(is.numeric), round, 4))

print("\nModel Performance Metrics:")
final_metrics

# Display diagnostic plots
final_diagnostics
```

# Conclusions & Limitations

How do sensor measurements (PT08.S1(CO) and PT08.S5(O3)) correlate with observed pollutant levels (NOx(GT) and CO(GT))?

Can these relationships improve the accuracy of carbon monoxide concentration predictions?



## Conclusions

The analysis directly addressed both research questions, revealing key insights about sensor correlations and prediction accuracy:

- Sensor pollutant correlations
  - CO sensor PT08.S1 showed strong correlation with CO(GT) levels
  - O3 sensor PT08.S5 showed significant correlation with CO levels, indicating important cross-sensitivity
  - NOx(GT) levels showed consistent positive correlation with CO(GT), though with a smaller magnitude
  - These correlations vary significantly by season, particularly the CO sensor, which shows reduced effectiveness in winter
  
- Prediction Accuracy Improvements
  - The inclusion of multiple sensor readings significantly improved prediction accuracy
  - Model comparison showed that incorporating both sensors & NOx(GT) greatly reduced RMSE (1.15 -> 0.603)
  - Final model achieved R^2 of 0.771, indicating that around 77% of CO variation can be explained using these sensor relationships
  - Seasonal interactions proved crucial, with the model capturing how sensor performance varise across different times of year
  

## Limitations

- Data limitations
  - Missing values & sensor failures required extensive cleaning
  - GT measurements may contain their own uncertainties
  - Dataset may not capture extreme events

- Modeling limitations
 - Linear model assumptions may oversimplify complex interactions in atmosphere
 - The unexplained 23% variation suggests other imporant factors may be missing
 - Residual analysis shows decreased accuracy at higher pollution levels
 
- Practical implementation limitations
  - Real-time standardization would be needed 
  - Seasonal classification must be accurate
  - Multi-sensor dependence creates vulnerability to sensor failures
