Name: Aditya Chauhan

ID: 169027493

Class: DATA 100 A

Professor: Dr. Devon Becker

Dataset: Air Quality (UC Irvine) (Vito, S. (2008). Air Quality [Dataset]. UCI Machine Learning Repository. <https://doi.org/10.24432/C59K5F>.)

# Introduction

Accurate air quality monitoring relies on sensor data to track pollutants like ozone, carbon monoxide, and nitrogen oxides. Understanding the relationship between sensor data and measured pollutants is crucial for validating monitoring technologies and enhancing predictive capabilities.

# Research Question: 

How do sensor measurements (e.g., PT08.S1(CO) and PT08.S5(O3)) correlate with observed pollutant levels such as NOx and CO, and can these correlations help refine air quality predictions?

Descriptive Features:
- Continuous
  - PT08.S1(CO)
  - PT08.S5(O3)
  - NOx(GT)
- Categorical
  - Season
  
Target Feature:
- CO(GT)

# Preprocessing

## Library Loads & Data Imports
```{r}
library(readxl)
library(tidyverse)
library(ggplot2)
library(tidymodels)
library(patchwork)
library(lubridate)
library(zoo)
library(reshape2)
library(dplyr)
library(RColorBrewer)

air_quality_data <- read_excel("AirQualityUCI.xlsx", sheet = "AirQualityUCI")
```

## Creating Categorical Features
```{r}
# Define pollutant columns based on existing column names
pollutant_columns <- c("CO(GT)", "NOx(GT)", "NO2(GT)")

# Grouping by Season and handling outliers in pollutant columns
air_quality_data <- air_quality_data |>
  group_by(Season = case_when(
    format(Date, "%m") %in% c("12", "01", "02") ~ "Winter",
    format(Date, "%m") %in% c("03", "04", "05") ~ "Spring",
    format(Date, "%m") %in% c("06", "07", "08") ~ "Summer",
    format(Date, "%m") %in% c("09", "10", "11") ~ "Autumn"
  )) |>
  mutate(
    across(all_of(pollutant_columns), 
           ~ ifelse(. > quantile(., 0.95, na.rm = TRUE), 
                    median(., na.rm = TRUE), .))
  ) |>
  ungroup()

# Create additional categorical features
air_quality_data <- air_quality_data |>
  mutate(
    DayOfWeek = weekdays(Date),
    
    Weekend = if_else(DayOfWeek %in% c("Saturday", "Sunday"), "Weekend", "Weekday"),
    
    Month = format(Date, "%B"), 
    
    Hour = as.numeric(format(Time, "%H")), 
    TimeOfDay = case_when(
      Hour >= 5 & Hour < 12 ~ "Morning",
      Hour >= 12 & Hour < 17 ~ "Afternoon",
      Hour >= 17 & Hour < 21 ~ "Evening",
      TRUE ~ "Night"
    ),
    
    # Temperature Categories
    TempCategory = case_when(
      T < 10 ~ "Cold",
      T >= 10 & T <= 25 ~ "Moderate",
      T > 25 ~ "Hot"
    ),
    
    # Humidity Levels
    HumidityLevel = case_when(
      RH < 30 ~ "Low",
      RH >= 30 & RH <= 60 ~ "Medium",
      RH > 60 ~ "High"
    )
  )

```

## Other Preprocessing

```{r}
# Handling missing values
air_quality_data <- air_quality_data |>
  mutate(
    across(c(`NOx(GT)`, `NO2(GT)`), ~ na_if(., -200))
  ) |>
  drop_na() 

# Normalize / Standardize Cont. Vars
air_quality_data <- air_quality_data |>
  mutate(
    across(c(T, RH, AH), ~ (. - mean(., na.rm = TRUE)) / sd(., na.rm = TRUE))
  )

# Interaction term
air_quality_data <- air_quality_data |>
  mutate(
    Temp_Humidity_Interaction = T * RH,
    CO_Temp_Interaction = `CO(GT)` * T
  )

# Rolling average (Sorted Chronologically)
air_quality_data <- air_quality_data |>
  arrange(Date, Time) |>
  mutate(
    Rolling_CO = rollmean(`CO(GT)`, k = 5, fill = NA)
  )

# Correlation Matrix
numeric_cols <- air_quality_data |>
  select(where(is.numeric))
cor_matrix <- cor(numeric_cols, use = "complete.obs")
#print(cor_matrix)

# 1. Correlation Matrix Plot
numeric_cols <- air_quality_data |>
  select(where(is.numeric))

cor_matrix <- cor(numeric_cols, use = "complete.obs")
cor_melt <- melt(cor_matrix)

plot1 <- ggplot(cor_melt, aes(x = Var1, y = Var2, fill = value)) +
  geom_tile(color = "white") + 
  scale_fill_gradient2(low = "blue", high = "red", mid = "white",
                       midpoint = 0, limit = c(-1, 1), space = "Lab",
                       name = "Correlation") +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1)
  ) + 
  labs(title = "Correlation Matrix", x = "Features", y = "Features")
# plot1

# Split dataset
set.seed(123)
data_split <- initial_split(air_quality_data, prop = 0.7)
train_data <- training(data_split)
test_data <- testing(data_split)

```


# Exploratory Analysis

```{r}

# Define dynamic limits
co_gt_limits <- c(0, 10)
ozone_sensor_limits <- c(200, 2500)
nox_gt_limits <- c(0, 500) 
sensor_limits <- c(0, 2000)

# 2. NOx(GT) vs CO(GT)
plot2 <- ggplot(air_quality_data, aes(x = `NOx(GT)`, y = `CO(GT)`)) +
  geom_point(alpha = 0.6, color = "steelblue", size = 1) +
  geom_smooth(method = "lm", se = TRUE, color = "darkred", linetype = "dashed", size = 1) +
  scale_y_continuous(limits = co_gt_limits) +
  scale_x_continuous(limits = nox_gt_limits) +
  labs(title = "NOx(GT) vs CO(GT)",
       x = "NOx(GT) (Nitrogen Oxides)",
       y = "CO(GT) (Carbon Monoxide)") +
  theme_minimal(base_size = 12) +
  theme(plot.title = element_text(face = "bold", size = 14))

# 3. Sensor Data vs CO(GT)
plot3 <- ggplot(air_quality_data, aes(x = `PT08.S1(CO)`, y = `CO(GT)`)) +
  geom_point(alpha = 0.6, aes(color = `PT08.S1(CO)`), size = 1.5) +
  geom_smooth(method = "lm", se = TRUE, color = "darkgreen", linetype = "solid", size = 1) +
  scale_color_gradient(low = "yellow", high = "red") +
  scale_y_continuous(limits = co_gt_limits) +
  scale_x_continuous(limits = sensor_limits) +
  labs(title = "Sensor Data vs CO(GT)",
       x = "PT08.S1(CO) (Sensor Data)",
       y = "CO(GT) (Carbon Monoxide)",
       color = "Sensor Value") +
  theme_minimal(base_size = 12) +
  theme(plot.title = element_text(face = "bold", size = 14))

# 4. Seasonal Trends in CO(GT)
plot4 <- ggplot(air_quality_data, aes(x = Season, y = `CO(GT)`, fill = Season)) +
  geom_boxplot(outlier.shape = 21, outlier.fill = "red", outlier.size = 2, alpha = 0.7) +
  scale_fill_brewer(palette = "Pastel2") +
  scale_y_continuous(limits = co_gt_limits) +
  labs(title = "Seasonal Trends in CO(GT)",
       x = "Season",
       y = "CO(GT) (Carbon Monoxide)") +
  theme_minimal(base_size = 12) +
  theme(plot.title = element_text(face = "bold", size = 14),
        legend.position = "none")

# 5. NOx(GT) vs Ozone Sensor Data
plot5 <- ggplot(air_quality_data, aes(x = `NOx(GT)`, y = `PT08.S5(O3)`)) +
  geom_point(alpha = 0.6, color = "cyan", size = 1.5) +
  geom_smooth(method = "lm", se = TRUE, color = "blue", size = 1.2, linetype = "solid") +
  scale_y_continuous(limits = c(0, 2500)) +
  scale_x_continuous(limits = nox_gt_limits) +
  labs(title = "NOx(GT) vs Ozone Sensor",
       x = "NOx(GT) (Nitrogen Oxides)",
       y = "PT08.S5(O3) (Ozone Sensor)") +
  theme_minimal(base_size = 12) +
  theme(plot.title = element_text(face = "bold", size = 14))

# Arrange plots with better spacing
final_plot <- (plot2 | plot3) / (plot4 | plot5) +
  plot_layout(guides = "collect") & theme(legend.position = "bottom")

final_plot

# Correlation Matrix for chosen features
correlation_data <- air_quality_data |>
  select(`NOx(GT)`, `CO(GT)`, `PT08.S1(CO)`, `PT08.S5(O3)`)
cor_matrix <- cor(correlation_data, use = "complete.obs")
# print(cor_matrix)

# Correlation Matrix Plot for chosen features
cor_melt <- melt(cor_matrix)
plot6 <- ggplot(cor_melt, aes(x = Var1, y = Var2, fill = value)) +
  geom_tile(color = "white") +
  scale_fill_gradient2(low = "blue", high = "red", mid = "white", 
                       midpoint = 0, limit = c(-1, 1), space = "Lab",
                       name = "Correlation") +
  theme_minimal(base_size = 12) +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1)) +
  labs(title = "Correlation Matrix: NOx(GT), CO(GT), Sensor Data, Ozone Sensor",
       x = "Features",
       y = "Features")

#plot6
```